Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-30 analysis/paper/paper.Rmd
TB Discussion edits

@@

@@

The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` and during the spring 2021 to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms), big.mark = ",")` which is `r round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. These densities for Deadwater Slough initially seem unrealistic especially given the size (volume) of the Salmon River at the slough relative to the Columbia River. However, we did catch `r prettyNum(nrow(cmr_df), big.mark = ",")` Northern Pikeminnow over the course of the study, of which only `r sum(cmr_df$capture_type == "Recapture", na.rm = T)` fish (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))`) were recaptures. Further, @Beamesderfer1991 did acknowledge that they were unsuccessful in finding a suitable method for sampling offshore; they noted that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective and so their abundance estimates only apply to nearshore areas. This was not the case at Deadwater Slough where maximum depth was approximately 10-15 m with a relatively homogeneous bottom contour i.e., angling was effective for the entire area. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977].

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers noticed that many were also large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality_sized" (larger) individuals and "stock-sized" individuals.

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Stock and quality size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020]. The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)`% of captured fish were over the quality size classification defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. In fact, the state record Northern Pikeminnow, which measured `r max(cmr_df$length, na.rm = T)` mm (`r round(max(cmr_df$length, na.rm = T) / 25.4, 1)` in) TL, was caught in Deadwater Slough during the fall 2020 effort.

@@

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Estimators assumed: (1) the population is closed (no immigration, emigration, births, or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) samples(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken. If the population is actually open (potential immigration and/or emigration during the survey period), our estimators should still provide unbiased estimates of abundance as long as the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), we are comfortable with the assumption of no immigration or emigration. Assumption (3) indicates whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that only affects the interpretation of the abundance estimates (i.e., our estimates are the abundance of Northern Pikeminnow susceptible to angling above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once; however, we did recapture two individual fish more than once suggesting that marked fish were still susceptible to angling. We are also comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks) given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.

We do consider two potential violations of assumption, both which reduces the probability that a marked fish is susceptible to capture in the second (and subsequent) samples. First, the assumption of a closed population could be violated if some marked fish that are released back to the population die. We don't believe this case to be likely, as fish that were dead or nearly so after processing were not released back to the population (nor recorded as such), and those that were released were hand-released only if they were able to swim away on their own volition; however, the potential for marked and released fish to have died at a higher rate than unmarked fish should be acknowledged. The second, and more likely, reason is that marked and released fish did not mix thoroughly back into the population (at least not at a fast enough rate), reducing their probability of capture in subsequent captures, thus violating assumption (2). We attempted to release marked fish throughout Deadwater Slough to facilitate mixing back into the population; however, anecdotally volunteer anglers would congregate to particular areas of the slough either to socialize with other anglers, because the fishing was good, etc. and if marked and released fish were not readily available in those areas that would make marked fish less susceptible to capture. This violation would, in effect, reduce the number of marked fish available for recapture or reduce the marking fraction in some areas of the slough. Violations of assumption (2) would lead to an overestimate of abundance of Northern Pikeminnow which is a real possibility in our study.

The spring abundance estimate was smaller than either fall abundance estimate. However, our approach assumed equal capture probabilities between the fall and spring sample efforts. If it does differ, we suspect that the capture probability in the spring would be lower due to higher flows (i.e., reduced densities), which would lead to our spring abundance estimate being biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 effort to be a conservative estimate.

@@

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. Aside from juvenile Chinook Salmon, identified species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing, [@Hodgson1998]), and further, angled fish are much more likely to have empty stomachs [@Jurajda2016]. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our results suggest. 
<!--
Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...
-->

@@

An average-size Northern Pikeminnow from Deadwater Slough is estimated to consume `r median(fall_g_summ$cumm_g)` g (range `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) and `r median(spring_g_summ$cumm_g)` g (range `r min(spring_g_summ$cumm_g)` - `r max(spring_g_summ$cumm_g)` g) of fish during spring and fall, respectively, to maintain its body weight using mean daily water temperatures recorded at a nearby stream gage [@USGS2016]. This equates to `r median(fall_g_summ$cumm_g) + median(spring_g_summ$cumm_g)` g of fish consumed during the DSR and NRR peak emigrations assuming the median value among diet scenarios (i.e., if 50% of their diet consists of fish); values ranged from a minimum of `r min(fall_g_summ$cumm_g) + min(spring_g_summ$cumm_g)` g if only 10% of the diet is fish to a maximum of `r max(fall_g_summ$cumm_g) + max(spring_g_summ$cumm_g)` g if 90% of the diet is fish.

<!-- The following needs to be updated after we finalize some results -->
This approach currently assumes that juvenile Chinook Salmon are the only fish prey available to or preyed upon by Northern Pikeminnow. It further assumes that all prey available have energy densities equal to juvenile Chinook Salmon, which may not be far from the truth, given that other salmonids including juvenile steelhead and Sockeye Salmon are likely also prevalent as available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. However, our study is focused on potential impacts to natural-origin DSR and NRR Chinook Salmon, in which case, let’s consider a “thought exercise” where DSR and NRR Chinook Salmon represent only 10% and 25% of total fish prey consumed by Northern Pikeminnow during these periods. Under these scenarios, the average Northern Pikeminnow would consume 8.95 g (range 4.52 – 10.05 g) if 10% of prey were natural-origin Chinook Salmon and 22.38 g (range 11.31 – 25.12) if 25%.

To estimate the total number of possible juvenile Chinook Salmon that could be consumed, we divided the cumulative consumption in grams for an individual Northern Pikeminnow by the average weight of a Chinook Salmon emigrant (taken from all rotary screw traps in the Upper Salmon MPG) to come up with a juvenile equivalency, and then multiplied this by a "stable" estimate of abundance for Northern Pikeminnow in Deadwater Slough, resulting in an estimate of the equivalent toal number of Chinook Salmon juveniles consumed. If we assume an average juvenile Chinook Salmon of 10.6 g, the mean from DSR and NRR Chinook Salmon emigrants in the Upper Salmon MPG, this equates to 0.84 juvenile Chinook Salmon (range 0.43 – 0.94) under the 10% scenario and 2.11 juvenile Chinook Salmon (range 1.07 – 2.37) under the 25% scenario.

Let’s now consider a stable Northern Pikeminnow population size of 10,000 (roughly the mean of fall and spring abundance estimates, divided by two to account for half of marked individuals not being available for recapture). In this case, the Northern Pikeminnow population in Deadwater Slough would be estimated to consume 8,400 (range 4,300 – 9,400) or 21,100 (range 10,700 – 23,700) under the 10% and 25% scenarios, respectively.

@@

<!-- Also update this after finalizing results -->
Finally, let's take our thought exercise one additional step and estimate the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. Our approach here was to multiply the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon. Assuming an SAR of `r round(sar$medsar, 5)` (SE: `r prettyNum(round(sar$sdsar, 5))`) we'll consider the "adult equivalents" to our estimates of juveniles consumed. Under this scenario, the consumed juveniles would be equivalent to 51.6 adults (range 26.4 – 57.7) and 129.6 adults (range 65.7 – 145.6). However, we believe these 10% and 25% scenarios to be conservative, especially during the fall DSR emigration period. During spring, natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, and Sockeye Salmon) are present in the river, and so 10% or 25% may be reasonable. However, during fall, DSR Chinook Salmon are likely the dominant prey species available to Chinook Salmon in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than in Chinook Salmon. Note that this thought exercise only includes the Northern Pikeminnow population in Deadwater Slough even though they are present throughout the upper Salmon River, although likely at lower densities than in Deadwater Slough, and also ignores potential impacts to ESA-listed steelhead and Sockeye Salmon that are also a portion of the Northern Pikeminnow diet.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile Chinook Salmon and other ESA-listed salmonids in the Deadwater Slough. the Deadwater Slough is recognized as an important bird watching and nesting area due to the associated upstream braided channels and riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using Deadwater Slough including the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the initial sampling effort in fall 2019, a two-person crew walked the entire reach including surrounding and upstream areas scanning for passive integrated transponder (PIT) tags that are implanted into thousands of natural- and hatchery-origin salmonids in the Upper Salmon River, annually, to monitor movement and survival. During that informal survey, 10 PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (4), and steelhead (3), all juveniles. Two of the Chinook Salmon and one steelhead were natural-origin, the remaining were of hatchery-origin. Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake and Columbia rivers, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although we did observe Belted Kingfishers, mergansers, and Great Blue Herons at Deadwater Slough during field efforts; Great Blue Herons were especially prevalent in the anastomizing reach immediately upriver of the slackwater reach. Additionally, there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, although the site is not within their breeding range; rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions and associated habitat features at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon and other ESA-listed salmonids from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (citation). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local business rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require year of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult

>>>>

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers reported that many were large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality_sized" (larger) individuals and "stock-sized" individuals.

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Stock and quality size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020].

@@

The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)`% of captured fish were over the quality size classification defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. In fact, the state record Northern Pikeminnow, which measured `r max(cmr_df$length, na.rm = T)` mm (`r round(max(cmr_df$length, na.rm = T) / 25.4, 1)` in) TL, was caught in Deadwater Slough during the fall 2020 effort.

@@

The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` and during the spring 2021 to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms), big.mark = ",")` which is `r round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater, respectively, than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Although estimated densities for Deadwater Slough are high given the relative size (volume) of the Salmon River at the slough relative to the Columbia River, high CPUE combined with low recapture rates (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))` fish were recaptures) suggest this area may support exceptionally high predator densities. Further, @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective. Conversely, at Deadwater Slough the maximum depth was approsimately 10-15 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere in the Columbia River basin. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977]. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin. @Widener2021 has identified a need for more current information on piscine predator populations in the Columbia River.

@@

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Yet, angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicting that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> We also recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would make marked fish less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability.

<!--
All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Estimators assumed: (1) the population is closed (no immigration, emigration, births, or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) samples(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken. If the population is actually open (potential immigration and/or emigration during the survey period), our estimators should still provide unbiased estimates of abundance as long as the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), we are comfortable with the assumption of no immigration or emigration. Assumption (3) indicates whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that only affects the interpretation of the abundance estimates (i.e., our estimates are the abundance of Northern Pikeminnow susceptible to angling above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once; however, we did recapture two individual fish more than once suggesting that marked fish were still susceptible to angling. We are also comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks) given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.

We do consider two potential violations of assumption, both which reduces the probability that a marked fish is susceptible to capture in the second (and subsequent) samples. First, the assumption of a closed population could be violated if some marked fish that are released back to the population die. We don't believe this case to be likely, as fish that were dead or nearly so after processing were not released back to the population (nor recorded as such), and those that were released were hand-released only if they were able to swim away on their own volition; however, the potential for marked and released fish to have died at a higher rate than unmarked fish should be acknowledged. The second, and more likely, reason is that marked and released fish did not mix thoroughly back into the population (at least not at a fast enough rate), reducing their probability of capture in subsequent captures, thus violating assumption (2). We attempted to release marked fish throughout Deadwater Slough to facilitate mixing back into the population; however, anecdotally volunteer anglers would congregate to particular areas of the slough either to socialize with other anglers, because the fishing was good, etc. and if marked and released fish were not readily available in those areas that would make marked fish less susceptible to capture. This violation would, in effect, reduce the number of marked fish available for recapture or reduce the marking fraction in some areas of the slough. Violations of assumption (2) would lead to an overestimate of abundance of Northern Pikeminnow which is a real possibility in our study.
-->

The spring abundance estimate was smaller than fall abundance estimates. Our approach assumed equal capture probabilities between the fall and spring sample efforts. Yet, we suspect that the capture probability in the spring could be lower as higher spring flows reduce Northern Pikeminnow densities biasing our spring abundance estimate low. Therefore, we believe our spring abundance estimate to be conservative.

@@

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. In addition to juvenile Chinook Salmon, species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing, [@Hodgson1998]), and further, angled fish are much more likely to have empty stomachs [@Jurajda2016]. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our results suggest. 
<!--
Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...
-->

@@

An average-size Northern Pikeminnow from Deadwater Slough is estimated to consume `r median(fall_g_summ$cumm_g)` g (range `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) and `r median(spring_g_summ$cumm_g)` g (range `r min(spring_g_summ$cumm_g)` - `r max(spring_g_summ$cumm_g)` g) of fish during spring and fall, respectively, to maintain its body weight. This equates to `r median(fall_g_summ$cumm_g) + median(spring_g_summ$cumm_g)` g of fish consumed during the DSR and NRR peak emigrations, assuming the median value among diet scenarios (i.e., if 50% of their diet consists of fish). Consumption ranged from `r min(fall_g_summ$cumm_g) + min(spring_g_summ$cumm_g)` for a diet comprised of 10% fish to `r max(fall_g_summ$cumm_g) + max(spring_g_summ$cumm_g)` for a diet comprised of 90% fish.

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Other salmonid including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. Other studies have used a similar approach where the energy densities of different salmonids were generalized using an estimate from a single species [@Peterson1999].<!-- The following needs to be updated after we finalize some results --> However, our study is focused on potential impacts to natural-origin DSR and NRR Chinook Salmon, in which case, let's consider a thought exercise where DSR and NRR Chinook Salmon represent only 10% and 25% of total fish prey consumed by Northern Pikemnnow during these periods. Under these scenarios, the average Northern Pikeminnow would consume 8.95 g (range 4.52 – 10.05 g) if 10% of prey were natural-origin Chinook Salmon and 22.38 g (range 11.31 – 25.12) if 25%.

Our estimated consumption of Chinook Salmon juveniles by Northern Pikeminnow was calculated using a static predator population. If we assume an average juvenile Upper Salmon MPG Chinook Salmon of 10.6 g, this equates to 0.84 juvenile Chinook Salmon (range 0.43 – 0.94) under the 10% scenario and 2.11 juvenile Chinook Salmon (range 1.07 – 2.37) under the 25% scenario. Let’s now consider a stable Northern Pikeminnow population size of 10,000 (roughly the mean of fall and spring abundance estimates, divided by two to account for half of marked individuals not being available for recapture). In this case, the Northern Pikeminnow population in Deadwater Slough would be estimated to consume 8,400 (range 4,300 – 9,400) or 21,100 (range 10,700 – 23,700) under the 10% and 25% scenarios, respectively.

@@

<!-- Also update this after finalizing results -->
Our primary objective was to assess the impact of predation on juvenile Chinook Salmon in the Upper Salmon MPG. However, there is also interest in estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. Our approach here was to multiply the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon. Assuming an SAR of `r round(sar$medsar, 5)` (SE: `r prettyNum(round(sar$sdsar, 5))`) we'll consider the "adult equivalents" to our estimates of juveniles consumed. Under this scenario, the consumed juveniles would be equivalent to 51.6 adults (range 26.4 – 57.7) and 129.6 adults (range 65.7 – 145.6). However, we believe these 10% and 25% scenarios to be conservative, especially during the fall DSR emigration period. During spring, natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, and Sockeye Salmon) are present in the river, and so 10% or 25% may be reasonable. However, during fall, DSR Chinook Salmon are likely the dominant prey species available to Chinook Salmon in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than in Chinook Salmon. Note that this thought exercise only includes the Northern Pikeminnow population in Deadwater Slough even though they are present throughout the upper Salmon River, although likely at lower densities than in Deadwater Slough, and also ignores potential impacts to ESA-listed steelhead and Sockeye Salmon that are also a portion of the Northern Pikeminnow diet.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile Chinook Salmon and other ESA-listed salmonids in the Deadwater Slough. the Deadwater Slough is recognized as an important bird watching and nesting area due to the associated upstream braided channels and riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using Deadwater Slough including the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the initial sampling effort in fall 2019, a two-person crew walked the entire reach including surrounding and upstream areas scanning for passive integrated transponder (PIT) tags that are implanted into thousands of natural- and hatchery-origin salmonids in the Upper Salmon River, annually, to monitor movement and survival. During that informal survey, 10 PIT tags were recovered near active bird nests and in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (4), and steelhead (3). Two of the Chinook Salmon and one steelhead were natural-origin, the remaining were of hatchery-origin. Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake and Columbia rivers, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although we did observe Belted Kingfishers, mergansers, and Great Blue Herons at Deadwater Slough during field efforts; Great Blue Herons were especially prevalent in the anastomizing reach immediately upriver of the slackwater reach. Additionally, there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough; however, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions and associated habitat features at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon and other ESA-listed salmonids from the many piscivorous birds known to use the site. Future studies of predation would benefit from estimates from piscivorous avian predators.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (citation). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local businesses rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require year of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult
